---
title: "GWAS2023analysis"
author: "Venkata Naresh Boddepalli"
date: "2023-12-24"
output:
  word_document: default
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
#install.packages("mvtnorm")

library(inti)
library(lme4)
library(GAPIT)

```



```{r}
getwd()
#list.dirs(".")
```

```{r}
#read data
pheno <- seedsize2224

#head(leaf_df)
#length(unique(pheno$geno))

##set as factors
cols <- c("block", "line", "env")

pheno[cols] <- lapply(pheno[cols], as.factor)
# After setting factors
pheno[cols] <- lapply(pheno[cols], as.factor)

# Convert all trait columns to numeric
pheno[4:ncol(pheno)] <- lapply(pheno[4:ncol(pheno)], function(x) as.numeric(as.character(x)))


###add total leaflet area
str(pheno)
```
##Model selection
```{r}
library(readxl)
library(MuMIn)
library(emmeans)

seed_A <- lmer(seed_area ~ env + (1|env:block) + line + (1|env:line), data = pheno)
seed_L <- lmer(seed_length ~ env + (1|env:block) + line + (1|env:line), data = pheno)
seed_W <- lmer(seed_width ~ env + (1|env:block) + line + (1|env:line), data = pheno)
seed_A.emm <- emmeans(seed_A, "line")
seed_L.emm <- emmeans(seed_L, "line")
seed_W.emm <- emmeans(seed_W, "line")
seed_A_blue <- as.data.frame(seed_A.emm)
seed_L_blue <- as.data.frame(seed_L.emm)
seed_W_blue <- as.data.frame(seed_W.emm)
write_xlsx(seed_A_blue, "seed_A.blue.xlsx")
write_xlsx(seed_L_blue, "seed_L.blue.xlsx")
write_xlsx(seed_W_blue, "seed_W.blue.xlsx")


```


```{r}
library(emmeans)
seed.emm <- emmeans(seed, "line")

seed_blue<- as.data.frame(seed.emm)

```

```{r}
#install.packages("writexl")
write_xlsx(seed_blue, "seed_blue.xlsx")
```


```{r}
start_time <- Sys.time()
#create final blups dataframe
line <- unique(pheno$line)
final_data <- data.frame(line)

##Final descriptive table
descriptive_stats <- data.frame(character(0), 
                                numeric(0), 
                                numeric(0),
                                numeric(0),
                                numeric(0),
                                numeric(0),
                                numeric(0),
                                numeric(0),stringsAsFactors=FALSE)
  #assign column names
colnames(descriptive_stats) <- c("Trait","Mean","Median","Min","Max","SD","CV","H2.cullis")

#create the anova sort of table
grp <- c("line", "env", "block", "Residual" )
variance_table <- data.frame(grp)



##list of traits
traits <- colnames(pheno)[4:ncol(pheno)]

##loop through the traits

t =1 
for(t in 1:length(traits)){
 
    trait = traits[t]
  
    hr <- H2cal(data = pheno,
            trait = trait,
            gen.name = "line",
            rep.n = 6,
            env.n = 3,
            env.name = unique(pheno$env),
            random.model = "1 + (1|env)+ (1|env:block)+ (1|line) + (1|env:line) ",
            fixed.model = "0 + (1|env) + (1|env:block)+ line + (1|env:line)",
            emmeans = FALSE,
            plot_diag = FALSE,
            outliers.rm = TRUE)
    
    ####### blups
    blup <- hr$blups
   
    
    #add blups to table
    final_data <- merge(final_data, blup, by = "line")
    

     ###########descriptive
     Mn <- mean(as.numeric(unlist(blup[,2])))#mean
     Md <- median(as.numeric(unlist(blup[,2])))
     Min <- min(as.numeric(unlist(blup[,2])))
     Max <- max(as.numeric(unlist(blup[,2])))
     Sd <- sd(as.numeric(unlist(blup[,2])))
     CV <- Sd/Mn
     H2.cullis <- hr$tabsmr$H2.c
     
     #create a vector with variables
     to_add <- c(trait,Mn,Md,Min,Max, Sd, CV, H2.cullis)
     #add to the df
     descriptive_stats <- descriptive_stats %>% add_row(Trait = trait,
                                                   Mean = Mn,
                                                   Median = Md,
                                                   Min = Min,
                                                   Max = Max,
                                                   SD = Sd,
                                                   CV = CV,
                                                   H2.cullis = H2.cullis)

     
       #descriptive
    descriptive_stats[, -1] <- round(descriptive_stats[, -1], 2)#apply to whole table
    
    ############## variance components
    y <- as.data.frame(VarCorr(hr$model)) #extract variance components as a table

    #extract variance components (column 1, 4 and round to 2 d.p)
    y1 <- y[, c(1,4)] %>% mutate_if(is.numeric, ~round(., 2))
    y1$percent <- (y1[,2]/ sum(y1$vcov)*100)  #new column with %

    y1$percent <- round(y1$percent, 1) #round off to o d.p.

    names(y1)[2]<-paste0(trait,"_var")
    
    
    variance_table <- merge(variance_table, y1, by='grp')
    
    print(paste0("Done with: ", trait))
}

#get sum of variance table components
###add totals row
row_totals <- colSums(variance_table[,2:ncol(variance_table)])
variance_table[nrow(variance_table) + 1, ] <- c(a = "Totals", as.list(row_totals))

#save
write.csv(final_data,"C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\descriptive stats\\blups.random.csv", row.names = F)

write.csv( descriptive_stats, "C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\descriptive stats\\summary.random.csv", row.names = F)
write.csv(variance_table, "C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\descriptive stats\\variance_table.random.csv", row.names = F)

end_time <- Sys.time()
print(paste0("This code took long: ", end_time-start_time))
```

end of BLUPs generating 
```{r}
print(hr$model %>% summary())
```



```{r}
library(data.table)

start_time <- Sys.time()

# Add total leaflet area calculation if needed
# pheno$Total_Leaflet_Area <- pheno$SPP_I + pheno$SPP_M  # Example if you want to sum these

# Pre-allocate results storage
traits <- colnames(pheno)[4:ncol(pheno)]
# If adding Total_Leaflet_Area:
# traits <- c(traits, "Total_Leaflet_Area")

results <- lapply(traits, function(trait) {
  hr <- H2cal(
    data = pheno,
    trait = trait,
    gen.name = "line",
    rep.n = 6,
    env.n = 3,
    env.name = unique(pheno$env),
    random.model = "1 + env + (1|env:block) + (1|line:env)",
    fixed.model = "0 + env + (1|env:block) + line + (1|line:env)",
    emmeans = TRUE,
    plot_diag = FALSE,
    outliers.rm = TRUE
  )
  
  blue_means <- as.data.frame(emmeans(hr$model, specs = "line"))
  colnames(blue_means) <- c("line", "value")
  
  # Descriptive stats
  desc <- data.frame(
    Trait = trait,
    Mean = mean(blue_means$value, na.rm = TRUE),
    Median = median(blue_means$value, na.rm = TRUE),
    Min = min(blue_means$value, na.rm = TRUE),
    Max = max(blue_means$value, na.rm = TRUE),
    SD = sd(blue_means$value, na.rm = TRUE),
    CV = sd(blue_means$value) / mean(blue_means$value),
    H2.cullis = hr$tabsmr$H2.c
  )
  
  # Variance components
  y <- as.data.frame(VarCorr(hr$model))
  y <- y[y$grp != "line", ]
  y$percent <- round(y$vcov / sum(y$vcov) * 100, 1)
  
  list(blues = blue_means, desc = desc, var = y)
})

# Combine results
final_data <- Reduce(
  function(x, y) merge(x, y, by = "line", all = TRUE),
  lapply(results, function(x) setNames(x$blues, c("line", names(x$blues)[2])))
)

descriptive_stats <- do.call(rbind, lapply(results, `[[`, "desc"))

variance_table <- Reduce(
  function(x, y) merge(x, y, by = "grp", all = TRUE),
  lapply(results, function(x) x$var)
)

# Save results
fwrite(final_data, "blues.csv")
fwrite(descriptive_stats, "summary_blues.csv")
fwrite(variance_table, "variance_table_blues.csv")

end_time <- Sys.time()
print(end_time - start_time)
```





gives you data frames that have genotype data based on the phenotype data provided.
can use for input into tassel also. 
But this will help in extractiung common genotypes across both phenotypic data and genotypic data file. This helps in anlyzing common lines acrross all the locations and years.

#run this code for function 
#DO NOT change any code in this chunk

```{r}
subset_geno_pheno = function(SNP_data, pheno_data){
  
  geno_list = pheno_data$name[pheno_data$name %in% colnames(SNP_data)]
  noSNP = pheno_data$name[!(pheno_data$name  %in% colnames(SNP_data))]
  
  geno_df = SNP_data %>% select(c(1:11, all_of(geno_list)))
  
  Pheno_df = pheno_data %>% filter(name %in% geno_list)
  
  my_list = list("SNP_data" = geno_df, "Pheno_data" = Pheno_df, "genos_withSNP" = geno_list, "genos_no_SNP" = noSNP)
  
  return(my_list)
  
}

```

#run this code for function 
#DO NOT change any code in this chunk

will write pheno and geno data to tassel format from the above functon to a file directory of your choosing. 
```{r}
tassel_phenotype = function(pheno_file, base_file_name, genotype_data){
  n_traits = dim(pheno_file)[2]-1
  
  A = c("taxa", rep("data",n_traits))
  
  pheno_file_m  = as.matrix(pheno_file)
  
  pheno_file1 = rbind(pheno_file_m, colnames(pheno_file))
  pheno_file1 = rbind(pheno_file1, A)
  A = c("<Phenotype>",rep("",n_traits))
  pheno_file1 = rbind(pheno_file1,A)
  n_rows = dim(pheno_file1)[1]

  pheno_file1 = pheno_file1[c(n_rows,n_rows-1,n_rows-2,1:(n_rows-3)),]
  
  pheno_file1[3,1] = "Taxa"
  
  write.table(pheno_file1, paste(base_file_name, "_phenotype.txt") , sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE )
  
  write.table(genotype_data, paste(base_file_name, "_genotype.txt") , sep = "\t", row.names = FALSE, quote = FALSE, col.names = TRUE )
  
  
  geno_order = colnames(genotype_data [,12:ncol(genotype_data)])
  
  pheno_order = pheno_file1[4:nrow(pheno_file1),1]
  pheno_order = unname(pheno_order)
  
  my_list = list("pheno_file" = pheno_file1,  "geno_order = Pheno order"= identical(geno_order, pheno_order) , "pheno_order = geno_order" = identical(pheno_order, geno_order))
  
  return(my_list)

}

```

#How to read in the genotype file to R so that you can subset for needed genotypes
#Change the code accroding to your working directory and files 

```{r}

geno_file = read.table("C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\blups\\MungGnm7_updatedHapmap.hmp.txt", comment.char = "")
geno_file[1:10,1:10]
colnames(geno_file) = geno_file[1,]
geno_file = geno_file[-1,]
geno_file[1:10,1:20]
geno_file
```



#Change the code accroding to your working directory and files 

Sample part of the genotype file

```{r}

geno_subset = geno_file[,1:250]
pheno_file = read.csv("C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\blups\\blups.csv")
#pheno_file = BLUPS
A = colnames(pheno_file)
A[1] = "name"
colnames(pheno_file) = A
colnames(pheno_file)
pheno_file
```

```{r}
install.packages("tidyverse")
```

subsetting your genotype and phenotype file

#DO NOT change any code in this chunk
```{r}
data_subset = subset_geno_pheno(SNP_data = geno_file, pheno_data = pheno_file)
```
#DO NOT change any code in this chunk
```{r}
data_subset$SNP_data
data_subset$Pheno_data
data_subset$genos_withSNP
data_subset$genos_no_SNP
```
Format for tassell
#Only change your working directory 
```{r}
tassel_data = tassel_phenotype(pheno_file = data_subset$Pheno_data, "C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\blups\\blups.csv", genotype_data = data_subset$SNP_data)

tassel_data$pheno_file
tassel_data$`geno_order = Pheno order`
tassel_data$`pheno_order = geno_order`
```


#End of data formatting tutorial. 
#This will give two outputs 1. phenotype and 2. genotype. (common lines)





#GAPIT starts 
```{r}

#setwd("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\BLINK_MLM")
myY <- read.table("C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\blups\\blups.csv _phenotype.txt", head = TRUE)
myG <- read.table("C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\Seed image analysis\\blups\\blups.csv _genotype.hmp.txt", comment.char = "")

#myG <- read.table("mdp_genotype_test.hmp.txt", head = FALSE)
myGAPIT <- GAPIT(G=myG, output.numerical=TRUE)
myGD= myGAPIT$GD
myGM= myGAPIT$GM


#myGD=read.table("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\2023\\GAPIT.Genotype.Numerical.txt",head=T)
#myGM=read.table("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\2023\\GAPIT.Genotype.map.txt",head=T)
#Step 2: Run GAPIT
#setwd("C:\\Users\\naresh89\\Desktop\\My projects\\GWAS\\Analysis_2024\\2023\\BLINK_farmCPU")
myGAPIT <- GAPIT(
  Y=myY,
  GD=myGD,
  GM=myGM,
  PCA.total=3,
  SNP.MAF = 0.05,
  Multiple_analysis = TRUE,
  model=c("BLINK", "MLM", "farmCPU")
  )

```

```{r}
str(myG)
str(myY)
```




```{r}
# Load the necessary library
library(qqman)

# Load your GWAS data
gwas_results <-read.csv("C:\\Users\\naresh89\\Desktop\\My projects\\gwas\\analysis2024\\numberofseedperpod\\GAPIT.Association.GWAS_Results.BLINK.SPP.csv" ) 

significant_snps <- gwas_results[gwas_results$P < 0.05, "SNP"]

# Manhattan plot
png("manhattan_plot.png", width = 11, height = 8, units = "in", res = 600)
manhattan(gwas_results, chr = "Chr", bp = "Pos", snp = "SNP", p = "P.value", 
          col = c("blue4", "orange3"), ylim = c(0, 12),suggestiveline = -log10(1e-05), genomewideline = -log10(5e-08),
          highlight = significant_snps,  # Highlight significant SNPs
          annotatePval = 0.05,  # Annotate SNPs with P < 0.05
          main = "Number of seeds per pod")

```

